---
title: "Applying tidy principles to investigating chromatin composition and architecture"
author: Jacques Serizay^[jacquesserizay@gmail.com,[Institut Pasteur, Paris](https://research.pasteur.fr/en/team/spatial-regulation-of-genomes/)]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying tidy principles to investigating chromatin composition and architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Manipulating genomic ranges data

*Recommended time: 10 min*

##  Importing `GRanges` from files

The `BiocIO::import()` generic function lets one import `bed` files (or alike) 
into `GRanges` object in R. 

```{r}
library(GenomicRanges)

library(rtracklayer)

bedf <- system.file('extdata', 'S288C-borders.bed', package = 'Bioc2024tidyWorkshop', mustWork = TRUE)

import(bedf)
```

The tidy way of importing genomic ranges in R, however, is to read files as 
a `data.frame` (or `tibble`) and _coerce_ it as a `GRanges` once it's imported in 
memory. 

```{r}
library(tidyverse)

tib <- read_tsv(bedf, col_names = FALSE)

tib 

library(plyranges)

gr <- as_granges(tib, seqnames = X1, start = X2, end = X3)

gr
```

Note how refering to column names is done using [tidy evaluation](https://dplyr.tidyverse.org/articles/programming.html#data-masking). 

##  Manipulating `GRanges` with tidy verbs

With `plyranges`, [a number of tidy operations](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1597-8/tables/1) are readily available, e.g.: 

- `mutate()`
- `select()`
- `filter()`
- `group_by()` + `summarize()`
- ... 

Just like with `tidyverse` functions, these operations work well with the native 
`|>` pipe in R. 

```{r}
gr |> 
    mutate(score = runif(n())) |> 
    filter(score > 0.2) |> 
    mutate(round_score = round(score, digits = 1)) |> 
    group_by(round_score) |> 
    summarize(mean = mean(score))
```

But really, what `plyranges` excels at is providing direct access to "protected" 
variables, i.e. `seqnames`, `start`, `end`, `width`, `strand`, ...

```{r}
gr |> 
    mutate(
        seqnames = factor('XVI', levels(seqnames)), 
        width = 1, 
        strand = rep(c('-', '+'), n()/2)
    )
```

Finally, a number of operations for genomic arithmetics are provided by `plyranges`: 

```{r}
gr |> 
    anchor_center() |> 
    stretch(extend = -1000) |> 
    shift_upstream(250) |> 
    flank_upstream(100)
```

# Manipulating genomic interaction data

## What are `GInteractions`? 

*Recommended time: 15 min*

### Creating a `GInteractions` object from scratch

`GInteractions` class of object combines pairs of interacting `GRanges` together. It is defined by the `InteractionSet` package. 

Each side of the pair is referred to as "anchors" (e.g. `anchors1` and `anchors2` refer to left-hand and right-hand side of the genomic interaction). 

```{r}
library(InteractionSet)

gr1 <- GRanges("I:10-50")

gr2 <- GRanges("I:100-110")

GInteractions(anchor1 = gr1, anchor2 = gr2)
```

A `GInteractions` object can also be generated by specific a set of `regions` (as a `GRanges` object with multiple entries)
and a set of numerical indices linking the different regions by pairs. 

```{r}
GInteractions(anchor1 = c(1, 2, 3), anchor2 = c(1, 4, 5), regions = gr)
```

### Importing genomic interaction data from files

A `GInteractions` object is typically imported in R from a file storing 
genomic interactions, e.g. a `bedpe` file or a `pairs` file, using the 
`as_ginteractions()` coercing function from the `plyinteractions` package.

```{r}
bedpef <- system.file('extdata', 'S288C-loops.bedpe', package = 'Bioc2024tidyWorkshop', mustWork = TRUE)

tib <- read_tsv(bedpef, col_names = FALSE) 

tib

library(plyinteractions)

gi <- tib |> 
    as_ginteractions(
        seqnames1 = X1, start1 = X2, end1 = X3, 
        seqnames2 = X4, start2 = X5, end2 = X6
    )

gi
```

## Manipulating `GInteractions` the tidy way

*Recommended time: 15 min*

### Moving anchors around

Just like `GRanges`, one can mutate columns, including "protected" columns, 
including `seqnames{12}`, `start{12}`, `end{12}`, `strand{12}` and `width{12}`. 

```{r}
gi |> 
    mutate(
        seqnames1 = factor('XVI', levels(seqnames1)), 
        strand1 = '+', 
        start2 = end1, 
        width2 = width1 + 100, 
        score = runif(length(gi)), 
        is_cis = ifelse(seqnames1 == seqnames2, TRUE, FALSE)
    )
```

### Filtering interactions

Here again, the `filter()` function can operate on `GInteractions` objects 
to filter entries using a boolean condition. 

```{r}
gi |> filter(seqnames1 == 'I')

gi |> filter(seqnames2 == 'I')

gi |> 
    mutate(score = runif(length(gi))) |> 
    filter(seqnames2 == 'I', score > 0.2)
```

### Overlapping anchors

`join_overlap_left()` function defined in `plyranges` package has been adapted
to work with 

```{r}
centros <- system.file('extdata', 'S288C-centromeres.txt', package = 'Bioc2024tidyWorkshop', mustWork = TRUE) |> 
    read_tsv() |> 
    as_granges(seqnames = seqID) |> 
    anchor_center() |> 
    stretch(20000)

gi |> 
    join_overlap_left(centros) |> 
    filter(!is.na(patternName))
```

`pin_anchors{12}` functions can be used to restrict the overlap function to 
only one or the other anchor of each interaction.

```{r}
gi |> 
    pin_anchors1() |> 
    join_overlap_left(centros) |> 
    filter(!is.na(patternName))

gi |> 
    pin_anchors2() |> 
    join_overlap_left(centros) |> 
    filter(!is.na(patternName))
```

## Real-world use case: computing a P(s)

*Recommended time: 20 min*

### Importing data from pairs file

Let's first import pairs as a `tibble`, then coerce them into `GInterctions`.  
The `mESCs.pairs.gz` example dataset is provided by the workshop package. 

```{r}
pairsf <- system.file('extdata', 'mESCs.pairs.gz', package = 'Bioc2024tidyWorkshop', mustWork = TRUE)

pairs <- read_tsv(pairsf, col_names = FALSE, comment = "#") |> 
    set_names(c(
        "ID", "seqnames1", "start1", "seqnames2", "start2", "strand1", "strand2"
    )) |> 
    as_ginteractions(end1 = start1, end2 = start2, keep.extra.columns = TRUE)
```

### Counting interactions by strands

We can get insights in the short-scale genomic-distance-dependent frequency of pairs 
according to their strand combination. 

```{r}
df <- pairs |> 
    add_pairdist() |> 
    filter(pairdist < 2000) |> 
    group_by(strand1, strand2, pairdist) |> 
    count() 

ggplot(df, aes(x = pairdist, y = n, col = interaction(strand1, strand2))) + 
    geom_smooth() + 
    scale_y_log10()
```

### Plot P(s)

More generally, we can compute genome-wide genomic-distance-dependent interaction frequency, or P(s). 

```{r}
x <- 1.1^(1:200-1)
lmc <- coef(lm(c(1,1161443398)~c(x[1], x[200])))
bins_breaks <- unique(round(lmc[2]*x + lmc[1]))
bins_widths <- lead(bins_breaks) - bins_breaks

# Bin distances
df <- pairs |> 
    add_pairdist(colname = 's') |> 
    mutate(
        binned_s = bins_breaks[as.numeric(cut(s, bins_breaks))], 
        bin_width = bins_widths[as.numeric(cut(s, bins_breaks))]
    ) |> 
    group_by(binned_s, bin_width) |> 
    count(name = "n") |>
    as_tibble() |> 
    mutate(Ps = n / sum(n) / bin_width)

ggplot(df, aes(x = binned_s, y = Ps)) + 
    geom_line() + 
    scale_y_log10() +
    scale_x_log10() +
    annotation_logticks() + 
    labs(x = "Genomic distance", y = "Contact frequency") + 
    theme_bw()
```

-------------

# Manipulating coverage data

## Introduction to `tidyCoverage`

*Recommended time: 10 min*

### `CoverageExperiment` and `AggregatedCoverage` class

Both classes are built on top of `SummarizedExperiment` class of objects. 

```{r}
library(tidyCoverage)

data(ce)

data(ac)

ce

ac
```

### Creating a `CoverageExperiment` object from tracks and features

```{r}
tracks <- BigWigFileList(c(
    mnase = system.file("extdata", "MNase.bw", package = "tidyCoverage"), 
    cohesin = system.file("extdata", "Scc1.bw", package = "tidyCoverage")  
))
features <- GRangesList(
    TSSs = system.file("extdata", "TSSs.bed", package = "tidyCoverage") |> import(), 
    TTSs = system.file("extdata", "TTSs.bed", package = "Bioc2024tidyWorkshop") |> import()
)

ce <- CoverageExperiment(
    tracks = tracks, 
    features = features,
    width = 3000, 
    ignore.strand = FALSE
)

ce

assay(ce, 'coverage')

dim(assay(ce, 'coverage')[['TSSs', 'mnase']])
```

### Tidy coverage? That's right!

```{r}
library(tidySummarizedExperiment)

ce

ce |> filter(features == 'TSSs')

ce |> slice(2)

ce |> 
    filter(track == 'mnase') |> 
    select(features, n)
```

### `expand()` or `aggregate()`

```{r}
tib <- expand(ce)

tib

ac <- aggregate(ce)

ac
```

### Example: visualizing aggregated coverage

```{r}
ac |>
    ggplot(aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_grid(track ~ features, scales = "free") + 
    labs(x = 'Distance from genomic features', y = 'Signal coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

## Real-world use case: studying epigenomic landscape of reulatory elements

*Recommended time: 20 min*

### Fetch coverage data from ENCODE

```{r}
library(AnnotationHub)
library(BiocParallel)
ah <- AnnotationHub()
ids <- c('AH32207', 'AH35187')
names(ids) <- c('DNAse', 'H3K4me3')
bws <- bplapply(ids, function(.x) ah[[.x]] |> resource()) |> BigWigFileList()
names(bws) <- names(ids)
```

### Import DNase peaks from ENCODE

```{r}
features <- list(DNase = ah[['AH30077']] |> filter(zScore > 100))
```

### Generating coverage aggregates over DNAse peak

```{r}
ce <- CoverageExperiment(bws, features, width = 2000, window = 10)
```

```{r}
aggregate(ce) |> 
    ggplot(aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_wrap(~track) + 
    labs(x = 'Distance from DNAse peak', y = 'Signal') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

```{r}
ranges.order <- ce |> 
    expand() |> 
    filter(track == 'DNase', coord.scaled >= -100, coord.scaled <= 100) |> 
    summarize(sum = sum(coverage, na.rm = TRUE)) |> 
    arrange(desc(sum)) |>
    pull(ranges) |> 
    rev()

ce |> 
    expand() |> 
    mutate(ranges = factor(ranges, levels = ranges.order)) |> 
    ggplot(aes(x = coord.scaled, y = ranges, fill = coverage)) + 
    geom_tile() +
    labs(x = 'Distance from DNAse peak', y = 'Signal') + 
    facet_grid(~track) +
    scale_x_continuous(expand = c(0, 0)) + 
    scale_fill_distiller(palette = "OrRd", na.value = '#fff5ed', direction = 1, limits = c(5, 60), oob = scales::squish) +
    theme_bw() + 
    theme(
        legend.position = 'top', 
        axis.ticks.y = element_blank(), axis.text.y = element_blank()
    )
```
